Carrier phase and amplitude estimation for phase shift keying using pilots and data (Appendix)

### 2012-08-16 paper1.tex
added to papers.  There's some fiddly stuff to do.  Should probably think about running the simulations if I'm tied of proving stuff.

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $X$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{X}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{X}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = X$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is indepedent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]  
Observe that for any real number $\phi$, the distribution of $X$ and $e^{j\phi}X$ are the same, that is, the distribution is invariant to phase rotation.  If $\expect\abs{X} = \expect Z$ is finite, then $X$ has zero mean because
\begin{align*}
\expect X &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
&= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
&= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
\end{align*}

We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + X.
\]
The joint distribution of $R$ and $\Phi$ can be shown to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{\sqrt{r^2 - 2r\cos\phi + 1}}.
\]
The mean of $R e^{j\Phi}$ is equal to one because the mean of $X$ is zero.  So,
\begin{equation}\label{eq:expectRepartRphi}
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\end{equation}
where $\Re(\cdot)$ denotes the real part, and
\begin{equation}\label{eq:expectImpartRphi}
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\end{equation}
where $\Im(\cdot)$ denotes the imaginary part.  %The next lemma will be useful for the analysis of the least squares estimator.

\begin{lemma}\label{lem:gg2cont}
Let $X = Z e^{j\Theta}$ be a circularly symmetric complex random variable with pdf $\tfrac{1}{2\pi}f_Z(z)$.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + X$ and with joint pdf $f(r,\phi)$.  Then
\[
g(\phi) = \int_{0}^\infty r f(r, \phi) dr,
\]
and
\[
g_2(\phi) = \int_{0}^\infty r^2 f(r, \phi) dr.
\]
are continuous functions from $\reals$ to $\reals$.
\end{lemma}
\begin{IEEEproof}
BLERG
\end{IEEEproof}

@@

\begin{lemma}
Let $X = Z e^{j\Theta}$ be a circularly symmetric complex random variable with pdf $\tfrac{1}{2\pi}f_Z(z)$ such that $z^{-1} f_Z(z)$ is nonincreasing with $z$.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + X$.  Let,
\[
h_1(x) = \expect R \cos(x + \Phi) \;\;\; \text{and} \;\;\; h_2(x) =  \expect R \cos\sfracpart{ x + \Phi}.
\]
Then $\abs{h_1(x)}$ is uniquely maximised at $x=0$ over the interval $[-\pi,\pi]$ and $\abs{h_2(x)}$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
\end{lemma}
Before giving the proof we discuss the requirements of this lemma.  The requirement that $z^{-1}f_{Z}(z)$ be non increasing implies that the pdf of $X$ decreases as we move away from the origin. That is, the pdf of $X$ in rectangular coordinates is $p(x,y) = z^{-1}f_{Z}(z)$, where $z = \sqrt{x^2 + y^2}$, and this is required to be non increasing with $z$.  For example, the zero mean complex Gaussian distribution with independent real and imaginary parts satisfies this requirement.
\begin{IEEEproof}
BLERG
\end{IEEEproof}

@@

We now analyse the asymptotic distribution of $\sqrt{L} m_L$.  Let $T_L(\lambda) = \expect G_L(\lambda)$.  Using~\eqref{eq:rhoGLZ},
\begin{align*}
\sqrt{L} m_L &= \sqrt{L} \rho_0 \big( G_L(\hat{\lambda}_L) - G(0) \big) \\
&= \sqrt{L} \rho_0 \big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L) + T_L(\hat{\lambda}_L) - G(0) \big).
\end{align*}
Lemma~\ref{lem:empiricprocforrho} shows that 
\[
\sqrt{L}\big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L)  \big) = o_P(1) + X_L,
\]
where $X_L = \sqrt{L}\big( G_L(0) - T_L(0)  \big)$.  Lemma~\ref{lem:HLtoG} shows that
\[
%\sqrt{L}\big( T_L(\hat{\lambda}_L) - G(0) \big) = \sqrt{L} \hat{\lambda}_L \big( ? + o_P(1) \big).
\sqrt{L}\big( T_L(\hat{\lambda}_L) - G(0) \big) = o_P(1).
\]
It follows that
\[
%\sqrt{L} m_L = o_P(1) + X_L  + \hat{\lambda}_L \big( ? + o_P(1) \big).
\sqrt{L} m_L =  \rho_0 X_L + o_P(1). 
\]
Lemma~\ref{lem:XL} shows that the distrubtion of $X_L$ converges to the normal with zero mean and variance $p B_1 + d B_2$ as $L\rightarrow\infty$ where $B_1$ and $B_2$ are defined in the statement of Theorem~\ref{thm:normality}.  Thus, the distribution of $\sqrt{L} m_L$ converges to the normal with zero mean and variance $\rho_0^2(p B_1 + d B_2)$ as required.  Because $X_L$ does not depend on $\hat{\lambda}_L$, it follows that $\covar(X_L, \sqrt{L}\hat{\lambda}_L) = 0$, and so,
\[
\covar(\sqrt{L}\hat{m}_L, \sqrt{L}\hat{\lambda}_L) = \covar(\rho_0 X_L + o_P(1), \sqrt{L}\hat{\lambda}_L) \rightarrow 0
\]
as $L \rightarrow \infty$.  It remains to prove the lemmas that we have used.

>>>>

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $X$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{X}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{X}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = X$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is indepedent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]  
Observe that for any real number $\phi$, the pdf of $X$ and $e^{j\phi}X$ are the same, that is, the pdf is invariant to phase rotation.  If $\expect\abs{X} = \expect Z$ is finite, then $X$ has zero mean because
\begin{align*}
 \expect X &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
 &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
 &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
 \end{align*}

We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + X.
\]
The joint pdf of $R$ and $\Phi$ can be shown to be
\begin{equation}\label{eq:pdfRPhi}
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{\sqrt{r^2 - 2r\cos\phi + 1}}.
\end{equation}
The mean of $R e^{j\Phi}$ is equal to one because the mean of $X$ is zero.  So,
\begin{equation}\label{eq:expectRepartRphi}
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\end{equation}
where $\Re(\cdot)$ denotes the real part, and
\begin{equation}\label{eq:expectImpartRphi}
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\end{equation}
where $\Im(\cdot)$ denotes the imaginary part.  The next result will be useful for the analysis of the least squares estimator.  %Me ommit the proof, which is straightforward, but tedious.

\begin{proposition}\label{lem:gg2cont}
Let $X = Z e^{j\Theta}$ be a circularly symmetric complex random variable with pdf $\tfrac{1}{2\pi}f_Z(z)$ that is continuous on the positive reals and with $\expect \abs{X}^2 = \expect Z^2$ finite.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + X$ and with joint pdf $f(r,\phi)$.  Then
\[
g(\phi) = \int_{0}^\infty r f(r, \phi) dr,
\]
and
\[
g_2(\phi) = \int_{0}^\infty r^2 f(r, \phi) dr
\]
are continuous functions from $\reals$ to $\reals$.
\end{proposition}
\begin{IEEEproof}
From~\eqref{eq:pdfRPhi},
\[
g(\phi) = \int_{0}^\infty \frac{r^2 f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{\sqrt{r^2 - 2r\cos\phi + 1}} dr,
\]
and
\[
g_2(\phi) = \int_{0}^\infty \frac{r^3 f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{\sqrt{r^2 - 2r\cos\phi + 1}} dr.
\]
The continuity of these functions follows immediately from the continuity of $f_Z$.
\end{IEEEproof}

@@

\begin{proposition}
Let $X = Z e^{j\Theta}$ be a circularly symmetric complex random variable with pdf $\tfrac{1}{2\pi}f_Z(z)$ that is continuous and differentiable on the positive reals.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + X$.  Let,
\[
h_1(x) = \expect R \cos(x + \Phi) \;\;\; \text{and} \;\;\; h_2(x) =  \expect R \cos\sfracpart{x + \Phi}.
\]
If $z^{-1} f_Z(z)$ is nonincreasing with $z$, then $\abs{h_1(x)}$ is uniquely maximised at $x=0$ over the interval $[-\pi,\pi]$ and $\abs{h_2(x)}$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
\end{proposition}
Before giving the proof we discuss the requirements of this lemma.  The requirement that $z^{-1}f_{Z}(z)$ be non increasing implies that the pdf of $X$ decreases as we move away from the origin. That is, the pdf of $X$ in rectangular coordinates is $p(x,y) = z^{-1}f_{Z}(z)$, where $z = \sqrt{x^2 + y^2}$, and this is required to be non increasing with $z$.  For example, the zero mean complex Gaussian distribution with independent real and imaginary parts satisfies this requirement.
\begin{IEEEproof}
BLERG
\end{IEEEproof}

@@

We now analyse the asymptotic distribution of $\sqrt{L} m_L$.  Let $T_L(\lambda) = \expect G_L(\lambda)$.  Using~\eqref{eq:rhoGLZ},
\begin{align*}
\sqrt{L} m_L &= \sqrt{L} \rho_0 \big( G_L(\hat{\lambda}_L) - G(0) \big) \\
&= \sqrt{L} \rho_0 \big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L) + T_L(\hat{\lambda}_L) - G(0) \big).
\end{align*}
Lemma~\ref{lem:empiricprocforrho} shows that 
\[
\sqrt{L}\big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L)  \big) = o_P(1) + X_L,
\]
where $X_L = \sqrt{L}\big( G_L(0) - T_L(0)  \big)$.  Lemma~\ref{lem:HLtoG} shows that
\[
%\sqrt{L}\big( T_L(\hat{\lambda}_L) - G(0) \big) = \sqrt{L} \hat{\lambda}_L \big( ? + o_P(1) \big).
\sqrt{L}\big( T_L(\hat{\lambda}_L) - G(0) \big) = o_P(1).
\]
It follows that
\[
%\sqrt{L} m_L = o_P(1) + X_L  + \hat{\lambda}_L \big( ? + o_P(1) \big).
\sqrt{L} m_L =  \rho_0 X_L + o_P(1). 
\]
Lemma~\ref{lem:XL} shows that the distrubtion of $X_L$ converges to the normal with zero mean and variance $p B_1 + d B_2$ as $L\rightarrow\infty$ where $B_1$ and $B_2$ are defined in the statement of Theorem~\ref{thm:normality}.  Thus, the distribution of $\sqrt{L} m_L$ converges to the normal with zero mean and variance $\rho_0^2(p B_1 + d B_2)$ as required.  Because $X_L$ does not depend on $\hat{\lambda}_L$, it follows that $\covar(X_L, \sqrt{L}\hat{\lambda}_L) = 0$, and so,
\[
\covar(\sqrt{L}\hat{m}_L, \sqrt{L}\hat{\lambda}_L) = \covar(\rho_0 X_L + o_P(1), \sqrt{L}\hat{\lambda}_L) \rightarrow 0
\]
as $L \rightarrow \infty$.  It remains to prove the lemmas that we have used.  These lemmas are proved in~\cite{McKilliam_leastsqPSKpilotsdata_2012appendix}.

### 2012-08-16 paper2.tex
added to papers.  There's some fiddly stuff to do.  Should probably think about running the simulations if I'm tied of proving stuff.

@@

Here we compile lemmas required for the proof almost sure convergence (Theorem~\ref{thm:consistency}) and asymptotic normality (Theorem~\ref{thm:normality}) from the paper~\cite{McKilliam_leastsqPSKpilotsdata_2012}.  This document is not self contained and it is intended that the reader start at~\cite{McKilliam_leastsqPSKpilotsdata_2012}.  The equation reference numbers in this document start at (?).  All reference numbers less than (?) refer back to~\cite{McKilliam_leastsqPSKpilotsdata_2012}.  This document is separated into 3 sections.  Section~\ref{sec:proof-almost-sureappendix} contains those lemmas required for the proof of almost sure convergence (Theorem~\ref{thm:consistency}).  Section~\ref{sec:proof-asympt-normappendix} contains those lemmas required for the proof of asymptotic normality (Theorem~\ref{thm:normality}).  Section~\ref{sec:an-empirical-process} contains proofs for lemmas that arise during Section~\ref{sec:proof-asympt-normappendix}.  These involve the proof of~\emph{tightness} for certain empirical processes~\cite{Billingsley1999_convergence_of_probability_measures,Dudley_unif_central_lim_th_1999,Shorak_emp_proc_stat_2009}.

@@

\begin{lemma}
Both $d_1(\lambda, 0) \leq C_\epsilon L $ and $d_2(\lambda, 0) \leq D_\epsilon L$ for all $\lambda \in B_\epsilon$.
\end{lemma}
\begin{IEEEproof}
BLERG
\end{IEEEproof}

>>>>

%set this to the number of the last equation in the previous document.
\setcounter{equation}{20}
\newcounter{lasteqplusone}
\setcounter{lasteqplusone}{\value{equation}}
\addtocounter{lasteqplusone}{1}

@@

Here we compile lemmas required for the proof almost sure convergence (Theorem~\ref{thm:consistency}) and asymptotic normality (Theorem~\ref{thm:normality}) from the paper~\cite{McKilliam_leastsqPSKpilotsdata_2012}.  This document is not self contained and it is intended that the reader start at~\cite{McKilliam_leastsqPSKpilotsdata_2012}.  The equation reference numbers in this document start at (\arabic{lasteqplusone}).  All equation numbers less than (\arabic{lasteqplusone}) refer back to~\cite{McKilliam_leastsqPSKpilotsdata_2012}.  This document is separated into 3 sections.  Section~\ref{sec:proof-almost-sureappendix} contains those lemmas required for the proof of almost sure convergence (Theorem~\ref{thm:consistency}).  Section~\ref{sec:proof-asympt-normappendix} contains those lemmas required for the proof of asymptotic normality (Theorem~\ref{thm:normality}).  Section~\ref{sec:an-empirical-process} contains proofs for lemmas that arise during Section~\ref{sec:proof-asympt-normappendix}.  These involve the proof of~\emph{tightness} for certain empirical processes~\cite{Billingsley1999_convergence_of_probability_measures,Dudley_unif_central_lim_th_1999,Shorak_emp_proc_stat_2009}.

@@

\begin{lemma}
Both $d_1(\lambda, 0) \leq C_\epsilon L $ and $d_2(\lambda, 0) \leq D_\epsilon L$ for all $\lambda \in B_\epsilon$.
\end{lemma}
\begin{IEEEproof}
We prove the lemma for $d_2$.  The proof for $d_1$ is similar.  Observe that $d_2(\lambda, 0) = c_1 + c_2$ where 
 \[
 c_1 = \sum_{i \in P} R_i\sabs{\sin(\lambda + \Phi_i) - \sin(\Phi_i) }
 \]
 and 
 \[
 c_2 = \sum_{i \in D} R_i \sabs{\sin\fracpart{\lambda + \Phi_i} - \sin\fracpart{\Phi_i}}.
 \]
 As $\sin(\cdot)$ is Lipshitz continuous with constant $K=1$,
 \[
 \sabs{\sin(\lambda + \Phi_i) - \sin(\Phi_i)} \leq K\sabs{\lambda} = \sabs{\lambda} < \epsilon
 \]
 for all $i \in P \cup D$, and therefore $c_1 \leq \epsilon \sum_{i \in P} R_i$.  Let $S = \{ i \in D \mid I_\epsilon(\Phi_i) = 1\}$. Now $c_2 = c_3 + c_4$ where 
 \[
 c_3 = \sum_{i \in D, i \in S} R_i \sabs{\sin\fracpart{\lambda + \Phi_i} - \sin\fracpart{\Phi_i}}
 \]
 and
 \[
 c_4 =  \sum_{i \in D, i \notin S} R_i \sabs{\sin\fracpart{\lambda + \Phi_i} - \sin\fracpart{\Phi_i}}.
 \]
 If $i \notin S$ then $\round{\lambda + \Phi_i} = \round{\Phi_i}$ for all $\sabs{\lambda} < \epsilon$ and therefore
 \begin{align*}
 \vert\sin&\fracpart{\lambda + \Phi_i} - \sin\fracpart{\Phi_i}\vert \\
 &= \sabs{\sin(\lambda + \Phi_i - \round{\Phi_i}) - \sin(\Phi_i - \round{\Phi_i})} \\
 &\leq \sabs{\lambda} < \epsilon
 \end{align*}
and so, 
\[
c_4 \leq \epsilon \sum_{i \in D, i \notin S} R_i \leq \epsilon \sum_{i \in D} R_i.
\]
Otherwise, if $i \in S$, then $\sabs{\sin\fracpart{\lambda + \Phi_i} - \sin\fracpart{\Phi_i}} \leq 2$ because $\sin(\cdot) \leq 1$.  Thus 
\[
c_3 \leq 2 \sum_{i \in D, i \in C} R_i \leq 2 \sum_{i \in D} R_i I_\epsilon(\Phi_i).
\]
Now
\begin{align*}
d_2(\lambda, 0) &= c_1 + c_2 \\
&\leq c_1 + c_3 + c_4 \\
&\leq \epsilon \sum_{i \in P \cup D} R_i + 2 \sum_{i \in D} R_i I_\epsilon(\Phi_i)
\end{align*}
as required
\end{IEEEproof}
